The ANthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2023-12-15 paper.tex
finish integration of the methodology chapter

\todo{kh: finish transforming methodology into latex}
\section{Methodology for developing ANNO}
For the development of the ANNO ontology, we applied the onto-axiomatic method, a combination of the axiomatic method with a top-level ontology [Baumann2014, Herre2010].

The axiomatic method comprises principles for developing theories or formal knowledge bases, which aim at the foundation, systematisation and formalisation of a knowledge domain [Baumann2014, Herre2010].
When knowledge is systematised, a set of categories is considered primitive or basic.
Such categories are not explicitly defined, but implicitly described by axioms [Hilbert1918].
An example of a primitive category is “part”.
New notions can be introduced by explicit definitions based on the primitive or already defined categories [Herre2010].

The considered axioms differ in their degree of abstraction.
At the most general level of abstraction, they are provided by top-level ontologies, whose axioms and categories can be applied to most domains of the world.
The onto-axiomatic method combines the axiomatic method with a top-level ontology, which is used to create more specialised core and domain-specific ontologies [Baumann2014].
Possible ways of discovering axioms in empirical domains are generalisation based on single cases and idealisation [Baumann2011].

We built ANNO on the basis of an ontology development schema that includes three main steps: 1. Domain Specification, 2. Conceptualisation and 3. Axiomatisation [Herre2010], and used the General Formal Ontology (GFO) [Loebe2022, Burek2020, Herre2010] as a top-level ontology in the sense of the onto-axiomatic method.

\subsection{Domain specification}
As part of the domain specification, the anthropology experts conducted an extensive review of existing literature and ontologies, analysing and classifying the relevant information.
Together with ontologists, relevant use cases and competence questions\footnotemark{} [XD2016, MOMo2023], as well as views and classification principles of the objects in the anthropology domain [Herre2010] were discussed.
\footnotetext{queries that the ontology must be able to answer}
%
The following main use cases and competence questions (subitems, selected examples) were defined:
\begin{enumerate}
\item Annotation of anthropological models (representations of anatomical entities) using a controlled vocabulary
\begin{enumerate}
\item Query all bone types
\item Query all parts (types) of a specific bone (type)
\item Query all parts (types) of a specific bone compound (type)
\item Query all tooth types
\item Query all parts (types) of a specific tooth (type)
\item $\ldots$
\end{enumerate}
\item Specification of spatial relations between anatomical entities
\begin{enumerate}
\item Query all defined spatial relations
\item What is the relative anatomical location of an anatomical entity in relation to another anatomical entity?
\item On which anatomical structure lies a defined measurement point?
\item Between which measurement points lies a defined anatomical line?
\item $\ldots$
\end{enumerate}
\item Specification of phenotyping functions
\begin{enumerate}
\item Query all phenotype specifications
\item Query the phenotyping function for determining a particular phenotype
\item Which parameters/variables (e.g., measurement distances/angles/points) are required to determine a particular phenotype?
\item $\ldots$
\end{enumerate}
\end{enumerate}
Relevant domain objects were identified on the basis of the defined use cases.
These include bones, teeth, phenotypes, as well as spatial anatomical entities.
The objects are classified according to their type and part-whole relationship.
The main requirement for the ANNO core ontology to cover the intended use cases is that it adequately represents the basic anatomical and anthropological entities, the spatial relations between them and the phenotyping functions, while remaining relatively simple and compact.
This approach allows on the one hand an easy manageability of the ontology for domain experts and on the other hand its efficient integration into software (e.g., AW3D) according to the three-ontology method [...].
Following this method, the software only needs to implement access to the core categories (of the core or task ontology), while their subcategories (from the domain-specific ontology) are processed dynamically.
This means, among other things, that all relevant anatomical entities (bones, teeth, their parts and composite structures) must each be located in one subtree (i.e., must have a common superclass) rather than spread across the entire ontology.
Furthermore, clear relations must exist between the core entities so that the dependencies between them can be efficiently resolved by the software.
We decided to develop a new ontology because none of the existing ontologies could cover all defined requirements and integrating (extending and adapting) all the necessary parts from the other ontologies would have been much more complex (see section $\ldots$).

@@

During the conceptualisation phase [Here2010], the core concepts (categories, classes) and relations were introduced that form ANNOdc (domain-core ontology) (fig. 2).
The concepts were created by generalising and classifying the domain objects.
To answer the competence questions of the first use case, for example, a distinction must be made between whole bones, bone parts and composite bone structures.
Therefore, the corresponding concepts "Bone", "Bone part" and "Bone compound" were introduced.
Further concepts were defined to represent spatial relations and phenotypes.
Furthermore, we identified relations (e.g., partOf, boundaryOf or locationOf) relevant to capture axioms covering the defined use cases (fig 2).

@@

For the axiomatisation and formal foundation of ANNO we utilised GFO as a top-level ontology and reused its categories, relations, axioms and modules (as a kind of Ontology Design Patterns [ODP2005, XD2016, MOMo2023]).
We instantiated specifically the GFO modules “Material objects”, “Attributives” and “Space” [Burek2020, Loebe2021, Loebe2018] and adapted them according to the requirements of the use cases.
Domain concepts introduced in the conceptualisation phase were embedded in GFO, i.e., defined as subcategories (subclasses) of certain GFO categories (fig. 1).
The required relations between the concepts were then either adopted from GFO or derived (specialised) from the corresponding GFO relation.
For example, GFO specifies that material objects occupy space regions [Loebe2021].
We integrated this pattern in ANNOdc by deriving the class anno:AnatomicalStructure from gfo:MaterialObject and anno:AnatomicalSpace from gfo:SpaceRegion.
In this way, we were able to adopt the whole axiom including the "occupies" relation between the two ANNO classes.
In a similar way, we applied the GFO pattern "point region is boundary of line region is boundary of surface region is boundary of space region" [Baumann2016] to the ANNO categories anatomical point, line, surface and space.
According to the GFO relator [Loebe2018] pattern, we defined the ANNO class “Relative anatomical location” including the two roles: target (locationOf) and reference (relativeTo) anatomical entity (chapter $\ldots$).

Further axioms were introduced to precisely define certain categories.
In addition to an explicit textual definition of the category "Bone compound", for example, we introduced the following axiom:

@@

The resulting OWL 2 ontology was generated using SMOG [...] and validated by the HermiT reasoner [...] and SHACL shapes (chapter $\ldots$).
In both cases, the consistency of the ontology was proven.
As a proof of concept, we integrated the ontology into the AW3D software [...] and were able to successfully realise all three intended use cases.
Thereby, most of the core categories (such as “Bone”, “Bone part”, “Bone compound”, “Tooth”, “Tooth part”, “Phenotype”, “Relative anatomical location”, “Anatomical line” and “Anatomical point”) and relation types (such as hasPart, boundaryOf, locationOf and derivedFrom) of ANNOdc were utilised and their subcategories and specific relations were dynamically queried (according to the three-ontology method).

@@

\todo{kh: finish statistical table}

\todo{kh: add literature refs}
[Herre2010] \url{https://doi.org/10.1007/978-90-481-8847-5\_14}
[Baumann2014] \url{https://doi.org/10.3233/AO-140136}
[Baumann2011] \url{https://arxiv.org/abs/1110.5832v1}
[Baumann2016] \url{https://doi.org/10.3233/978-1-61499-660-6-53}
[Hilbert1918] Hilbert, D.(1918). Axiomatisches Denken. Mathematische Annalen, 78, 405–415. (in German)
[ODP2005] A. Gangemi: Ontology design patterns for semantic web content. Lecture Notes in Computer Science (2005)
[Burek2020] \url{https://doi.org/10.3233/FAIA200658}
[Loebe2021] \url{https://ceur-ws.org/Vol-2969/paper69-FOUST.pdf}
[Loebe2018] \url{https://ceur-ws.org/Vol-2205/paper29\_ontocom3.pdf}
[MOMo2023] \url{https://doi.org/10.3233/SW-222886}
[XD2016] \url{https://doi.org/10.3233/978-1-61499-676-7-23}
[Loebe2022] \url{https://doi.org/10.3233/AO-220264}

>>>>

\section{Methodology for developing ANNO}
For the development of ANNO, we applied the onto-axiomatic method, a combination of the axiomatic method with a top-level ontology \citep{baumann2014, herre2010}.
The axiomatic method comprises principles for developing theories or formal knowledge bases, which aim at the foundation, systematisation and formalisation of a knowledge domain \citep{baumann2014, herre2010}.
When knowledge is systematised, a set of categories is considered \emph{primitive} or \emph{basic}.
Such categories are not explicitly defined, but implicitly described by axioms \citep{Hilbert1918}.
An example of a primitive category is \enquote{part}.
New notions can be introduced by explicit definitions based on the primitive or already defined categories \citep{herre2010}.

The considered axioms differ in their degree of abstraction.
At the most general level of abstraction, they are provided by top-level ontologies, whose axioms and categories can be applied to most domains of the world.
The onto-axiomatic method combines the axiomatic method with a top-level ontology, which is used to create more specialised core and domain-specific ontologies \citep{baumann2014}.
Possible ways of discovering axioms in empirical domains are generalisation based on single cases and idealisation \citep{baumann2011}.

We built ANNO on the basis of an ontology development schema that includes three main steps: 1. Domain Specification, 2. Conceptualisation and 3. Axiomatisation \citep{herre2010}, and used the General Formal Ontology (GFO) \citep{Loebe2022, Burek2020, herre2010} as a top-level ontology in the sense of the onto-axiomatic method.

\subsection{Domain specification}
As part of the domain specification, the anthropology experts conducted an extensive review of existing literature and ontologies, analysing and classifying the relevant information.
Together with ontologists, relevant use cases and competence questions\footnotemark{} \citep{XD2016, MOMo2023}, as well as views and classification principles of the objects in the anthropology domain \citep{herre2010} were discussed.
\footnotetext{queries that the ontology must be able to answer}
%
The following main use cases and competence questions (subitems, selected examples) were defined:
\begin{enumerate}
\item Annotation of anthropological models (representations of anatomical entities) using a controlled vocabulary
\begin{enumerate}
\item Query all bone types
\item Query all parts (types) of a specific bone (type)
\item Query all parts (types) of a specific bone compound (type)
\item Query all tooth types
\item Query all parts (types) of a specific tooth (type)
\item $\ldots$
\end{enumerate}
\item Specification of spatial relations between anatomical entities
\begin{enumerate}
\item Query all defined spatial relations
\item What is the relative anatomical location of an anatomical entity in relation to another anatomical entity?
\item On which anatomical structure does a defined measurement point lie?
\item Between which measurement points does a defined anatomical line lie?
\item $\ldots$
\end{enumerate}
\item Specification of phenotyping functions
\begin{enumerate}
\item Query all phenotype specifications
\item Query the phenotyping function for determining a particular phenotype
\item Which parameters/variables (e.g., measurement distances/angles/points) are required to determine a particular phenotype?
\item $\ldots$
\end{enumerate}
\end{enumerate}
Relevant domain objects were identified on the basis of the defined use cases.
These include bones, teeth, phenotypes, as well as spatial anatomical entities.
The objects are classified according to their type and part-whole relationship.
The main requirement for the ANNO core ontology to cover the intended use cases is that it adequately represents the basic anatomical and anthropological entities, the spatial relations between them and the phenotyping functions, while remaining relatively simple and compact.
This approach allows on the one hand an easy manageability of the ontology for domain experts and on the other hand its efficient integration into software (e.g., AW3D) according to the three-ontology method [...].
Following this method, the software only needs to implement access to the core categories (of the core or task ontology), while their subcategories (from the domain-specific ontology) are processed dynamically.
This means, among other things, that all relevant anatomical entities (bones, teeth, their parts and composite structures) must each be located in one subtree (i.e., must have a common superclass) rather than spread across the entire ontology.
Furthermore, clear relations must exist between the core entities so that the dependencies between them can be efficiently resolved by the software.
We decided to develop a new ontology because none of the existing ontologies could cover all defined requirements and integrating (extending and adapting) all the necessary parts from the other ontologies would have been much more complex (see section $\ldots$).

@@

During the conceptualisation phase \citep{herre2010}, the core concepts (categories, classes) and relations were introduced that form ANNOdc (domain-core ontology) (fig. 2).
The concepts were created by generalising and classifying the domain objects.
To answer the competence questions of the first use case, for example, a distinction must be made between whole bones, bone parts and composite bone structures.
Therefore, the corresponding concepts \enquote{Bone}, \enquote{Bone part} and \enquote{Bone compound} were introduced.
Further concepts were defined to represent spatial relations and phenotypes.
Furthermore, we identified relations (e.g., partOf, boundaryOf or locationOf) relevant to capture axioms covering the defined use cases, see \cref{fig:core}.

@@

For the axiomatisation and formal foundation of ANNO we utilised GFO as a top-level ontology and reused its categories, relations, axioms and modules (as a kind of Ontology Design Patterns \citep{ODP2005, XD2016, MOMo2023}).
We instantiated specifically the GFO modules \enquote{Material objects}, \enquote{Attributives} and \enquote{Space} \citep{Burek2020, Loebe2021, Loebe2018} and adapted them according to the requirements of the use cases.
Domain concepts introduced in the conceptualisation phase were embedded in GFO, i.e., defined as subcategories (subclasses) of certain GFO categories (fig. 1).
The required relations between the concepts were then either adopted from GFO or derived (specialised) from the corresponding GFO relation.
For example, GFO specifies that material objects occupy space regions \citep{Loebe2021}.
We integrated this pattern in ANNOdc by deriving the class anno:AnatomicalStructure from gfo:MaterialObject and anno:AnatomicalSpace from gfo:SpaceRegion.
In this way, we were able to adopt the whole axiom including the \enquote{occupies} relation between the two ANNO classes.
In a similar way, we applied the GFO pattern \enquote{point region is boundary of line region is boundary of surface region is boundary of space region} \citep{baumann2016} to the ANNO categories anatomical point, line, surface and space.
According to the GFO relator \citep{Loebe2018} pattern, we defined the ANNO class \enquote{Relative anatomical location} including the two roles: target (locationOf) and reference (relativeTo) anatomical entity (chapter $\ldots$).

Further axioms were introduced to precisely define certain categories.
In addition to an explicit textual definition of the category \enquote{Bone compound}, for example, we introduced the following axiom:

@@

The resulting OWL 2 ontology was generated using SMOG [...] and validated by the HermiT reasoner [...] and SHACL shapes (chapter $\ldots$).
In both cases, the consistency of the ontology was proven.
As a proof of concept, we integrated the ontology into the AW3D software [...] and were able to successfully realise all three intended use cases.
Thereby, most of the core categories (such as \enquote{Bone}, \enquote{Bone part}, \enquote{Bone compound}, \enquote{Tooth}, \enquote{Tooth part}, \enquote{Phenotype}, \enquote{Relative anatomical location}, \enquote{Anatomical line} and \enquote{Anatomical point}) and relation types (such as hasPart, boundaryOf, locationOf and derivedFrom) of ANNOdc were utilised and their subcategories and specific relations were dynamically queried (according to the three-ontology method).

@@
